---
title: "Geospatial Carpentry | Session 2: Working with vector data"
author: "Claudiu Forgaci"
date: "2022-11-18"
output: html_document
---

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

```{r lib}
# Install packages
# library(remotes)
# remotes::install_github('ropensci/osmdata')

# Data manipulation
# library(tibble)
# library(dplyr)
# library(reshape)

# Data visualisation
library(ggplot2)

# Working with spatial data
library(sf)
# library(leaflet)
# library(raster)
# library(rgdal)
# library(proj4)
# library(scales)

# Working with OSM data
# library(osmdata)
# library(osmextract)
```

# Open and plot shapefiles (20 + 10 minutes)

## Import shapefiles

```{r}
aoi_boundary_HARV <- st_read(
  "data/NEON-DS-Site-Layout-Files/HARV/HarClip_UTMZ18.shp")
```

## Shapefile Metadata & Attributes

```{r}
st_geometry_type(aoi_boundary_HARV)
```

```{r}
st_crs(aoi_boundary_HARV)
```

```{r}
st_bbox(aoi_boundary_HARV)
```

```{r}
aoi_boundary_HARV
```

## Plot a shapefile

```{r}
# ggplot already introduced in the morning
ggplot() + 
  geom_sf(data = aoi_boundary_HARV, size = 3, color = "black", fill = "cyan1") + 
  ggtitle("AOI Boundary Plot") + 
  coord_sf()
```

```{r}
lines_HARV <- st_read("data/NEON-DS-Site-Layout-Files/HARV/HARV_roads.shp")
```

```{r}
point_HARV <- st_read("data/NEON-DS-Site-Layout-Files/HARV/HARVtower_UTM18N.shp")
```

```{r}
class(lines_HARV)
```

```{r}
class(point_HARV)
```

```{r}
st_crs(lines_HARV)
```

```{r}
st_bbox(lines_HARV)
```

```{r}
st_crs(point_HARV)
```

```{r}
st_bbox(point_HARV)
```

# Explore and plot by vector layer attributes (40 + 20 minutes)

```{r}
# data frames introduced in the morning; sf objects have additional metadata and a 'sticky' geometry column
point_HARV
```

```{r}
ncol(lines_HARV)
```

```{r}
names(lines_HARV)
```

```{r}
head(lines_HARV)
```

### Challenge

```{r}
ncol(point_HARV)
```

```{r}
ncol(aoi_boundary_HARV)
```

```{r}
# $ operator already introduced in the morning
point_HARV$Ownership
```

```{r}
names(point_HARV)
```

## Explore values within one attribute

```{r}
lines_HARV$TYPE
```

```{r}
levels(factor(lines_HARV$TYPE))
```

## Subset features

```{r}
# data wragling with dplyr introduced in the morning
footpath_HARV <- lines_HARV %>% 
  filter(TYPE == "footpath")
nrow(footpath_HARV)
```

```{r}
ggplot() + 
  geom_sf(data = footpath_HARV) +
  ggtitle("NEON Harvard Forest Field Site", subtitle = "Footpaths") + 
  coord_sf()
```

```{r}
ggplot() + 
  geom_sf(data = footpath_HARV, aes(color = factor(OBJECTID)), size = 1.5) +
  labs(color = 'Footpath ID') +
  ggtitle("NEON Harvard Forest Field Site", subtitle = "Footpaths") + 
  coord_sf()
```

### Challenge

```{r}
boardwalk_HARV <- lines_HARV %>% 
  filter(TYPE == "boardwalk")
```

```{r}
nrow(boardwalk_HARV)
```

```{r}
ggplot() + 
  geom_sf(data = boardwalk_HARV, size = 1.5) +
  ggtitle("NEON Harvard Forest Field Site", subtitle = "Boardwalks") + 
  coord_sf()
```

```{r}
stoneWall_HARV <- lines_HARV %>% 
  filter(TYPE == "stone wall")
nrow(stoneWall_HARV)
```

```{r}
ggplot() +
  geom_sf(data = stoneWall_HARV, aes(color = factor(OBJECTID)), size = 1.5) +
  labs(color = 'Wall ID') +
  ggtitle("NEON Harvard Forest Field Site", subtitle = "Stonewalls") + 
  coord_sf()
```

## Customize plots

```{r}
levels(factor(lines_HARV$TYPE))
```

```{r}
road_colors <- c("blue", "green", "navy", "purple")
```

```{r}
ggplot() +
  geom_sf(data = lines_HARV, aes(color = TYPE)) + 
  scale_color_manual(values = road_colors) +
  labs(color = 'Road Type') +
  ggtitle("NEON Harvard Forest Field Site", subtitle = "Roads & Trails") + 
  coord_sf()
```

## Adjust line width

```{r}
line_widths <- c(1, 2, 3, 4)
```

```{r}
ggplot() +
  geom_sf(data = lines_HARV, aes(color = TYPE, size = TYPE)) + 
  scale_color_manual(values = road_colors) +
  labs(color = 'Road Type') +
  scale_size_manual(values = line_widths) +
  ggtitle("NEON Harvard Forest Field Site", subtitle = "Roads & Trails - Line width varies") + 
  coord_sf()
```

### Challenge

```{r}
levels(factor(lines_HARV$TYPE))
```


```{r}
line_width <- c(1, 3, 2, 6)
```

```{r}
ggplot() +
  geom_sf(data = lines_HARV, aes(size = TYPE)) +
  scale_size_manual(values = line_width) +
  ggtitle("NEON Harvard Forest Field Site", subtitle = "Roads & Trails - Line width varies") + 
  coord_sf()
```

## Add plot legend

```{r}
ggplot() + 
  geom_sf(data = lines_HARV, aes(color = TYPE), size = 1.5) +
  scale_color_manual(values = road_colors) +
  labs(color = 'Road Type') + 
  ggtitle("NEON Harvard Forest Field Site", 
          subtitle = "Roads & Trails - Default Legend") + 
  coord_sf()
```

```{r}
ggplot() + 
  geom_sf(data = lines_HARV, aes(color = TYPE), size = 1.5) +
  scale_color_manual(values = road_colors) + 
  labs(color = 'Road Type') +
  theme(legend.text = element_text(size = 20), 
        legend.box.background = element_rect(size = 1)) + 
  ggtitle("NEON Harvard Forest Field Site", 
          subtitle = "Roads & Trails - Modified Legend") +
  coord_sf()
```

```{r}
new_colors <- c("springgreen", "blue", "magenta", "orange")

ggplot() + 
  geom_sf(data = lines_HARV, aes(color = TYPE), size = 1.5) + 
  scale_color_manual(values = new_colors) +
  labs(color = 'Road Type') +
  theme(legend.text = element_text(size = 20), 
        legend.box.background = element_rect(size = 1)) + 
  ggtitle("NEON Harvard Forest Field Site", 
          subtitle = "Roads & Trails - Pretty Colors") +
  coord_sf()
```

### Challenge 

```{r}
class(lines_HARV$BicyclesHo)
```

```{r}
levels(lines_HARV$BicyclesHo)
```

```{r}
lines_removeNA <- lines_HARV[!is.na(lines_HARV$BicyclesHo),] 
```

```{r}
# First, create a data frame with only those roads where bicycles and horses are allowed
lines_showHarv <- lines_removeNA %>% filter(BicyclesHo == "Bicycles and Horses Allowed")

# Next, visualise using ggplot
ggplot() + 
  geom_sf(data = lines_HARV) + 
  geom_sf(data = lines_showHarv, aes(color = BicyclesHo), size = 2) + 
  scale_color_manual(values = "magenta") +
  ggtitle("NEON Harvard Forest Field Site", subtitle = "Roads Where Bikes and Horses Are Allowed") + 
  coord_sf()
```

```{r}
state_boundary_US <- 
st_read("data/NEON-DS-Site-Layout-Files/US-Boundary-Layers/US-State-Boundaries-Census-2014.shp")
```

```{r}
levels(factor(state_boundary_US$region))
```

```{r}
colors <- c("purple", "springgreen", "yellow", "brown", "navy")
```

```{r}
ggplot() +
  geom_sf(data = state_boundary_US, aes(color = region), size = 1) +
  scale_color_manual(values = colors) +
  ggtitle("Contiguous U.S. State Boundaries") + 
  coord_sf()
```

# Plot multiple shapefiles (40 + 20 minutes - 15 minutes from plot costomisation)

```{r}
ggplot() + 
  geom_sf(data = aoi_boundary_HARV, fill = "grey", color = "grey") +
  geom_sf(data = lines_HARV, aes(color = TYPE), size = 1) +
  geom_sf(data = point_HARV) +
  ggtitle("NEON Harvard Forest Field Site") + 
  coord_sf()
```

```{r}
ggplot() + 
  geom_sf(data = aoi_boundary_HARV, fill = "grey", color = "grey") +
  geom_sf(data = lines_HARV, aes(color = TYPE),
          show.legend = "line", size = 1) +
  geom_sf(data = point_HARV, aes(fill = Sub_Type), color = "black") +
  scale_color_manual(values = road_colors) +
  scale_fill_manual(values = "black") +
  ggtitle("NEON Harvard Forest Field Site") + 
  coord_sf()
```

```{r}
ggplot() + 
  geom_sf(data = aoi_boundary_HARV, fill = "grey", color = "grey") +
  geom_sf(data = point_HARV, aes(fill = Sub_Type)) +
  geom_sf(data = lines_HARV, aes(color = TYPE), show.legend = "line",
          size = 1) + 
  scale_color_manual(values = road_colors, name = "Line Type") + 
  scale_fill_manual(values = "black", name = "Tower Location") + 
  ggtitle("NEON Harvard Forest Field Site") + 
  coord_sf()
```

```{r}
ggplot() +
  geom_sf(data = aoi_boundary_HARV, fill = "grey", color = "grey") +
  geom_sf(data = point_HARV, aes(fill = Sub_Type), shape = 15) +
  geom_sf(data = lines_HARV, aes(color = TYPE),
          show.legend = "line", size = 1) +
  scale_color_manual(values = road_colors, name = "Line Type") +
  scale_fill_manual(values = "black", name = "Tower Location") +
  ggtitle("NEON Harvard Forest Field Site") + 
  coord_sf()
```

### Challenge

```{r}
plot_locations <- st_read("data/NEON-DS-Site-Layout-Files/HARV/PlotLocations_HARV.shp")
```

```{r}
levels(factor(plot_locations$soilTypeOr))
```

```{r}
blue_orange <- c("cornflowerblue", "darkorange")
```

```{r}
ggplot() + 
  geom_sf(data = lines_HARV, aes(color = TYPE), show.legend = "line") + 
  geom_sf(data = plot_locations, aes(fill = soilTypeOr), 
          shape = 21, show.legend = 'point') + 
  scale_color_manual(name = "Line Type", values = road_colors,
     guide = guide_legend(override.aes = list(linetype = "solid", shape = NA))) + 
  scale_fill_manual(name = "Soil Type", values = blue_orange,
     guide = guide_legend(override.aes = list(linetype = "blank", shape = 21, colour = NA))) + 
  ggtitle("NEON Harvard Forest Field Site") + 
  coord_sf()
```

```{r}
ggplot() + 
  geom_sf(data = lines_HARV, aes(color = TYPE), show.legend = "line", size = 1) + 
  geom_sf(data = plot_locations, aes(fill = soilTypeOr, shape = soilTypeOr),
          show.legend = 'point', size = 3) + 
  scale_shape_manual(name = "Soil Type", values = c(21, 22)) +
  scale_color_manual(name = "Line Type", values = road_colors,
     guide = guide_legend(override.aes = list(linetype = "solid", shape = NA))) + 
  scale_fill_manual(name = "Soil Type", values = blue_orange,
     guide = guide_legend(override.aes = list(linetype = "blank", shape = c(21, 22),
     color = blue_orange))) + 
  ggtitle("NEON Harvard Forest Field Site") + 
  coord_sf()
```

```{r}
# we will come back to raster data on day 2
# DROP THIS CHALLENGE
ggplot() +
  geom_raster(data = CHM_HARV_df, aes(x = x, y = y, fill = layer)) +
  geom_sf(data = lines_HARV, color = "black") +
  geom_sf(data = aoi_boundary_HARV, color = "grey20", size = 1) +
  geom_sf(data = point_HARV, pch = 8) +
  ggtitle("NEON Harvard Forest Field Site w/ Canopy Height Model") + 
  coord_sf()
```

# Handling Spatial Projections and CRS (40 + 20 minutes)

## Working with spatial data from different sources

```{r}
state_boundary_US <- st_read("data/NEON-DS-Site-Layout-Files/US-Boundary-Layers/US-State-Boundaries-Census-2014.shp")
```

```{r}
ggplot() +
  geom_sf(data = state_boundary_US) +
  ggtitle("Map of Contiguous US State Boundaries") +
  coord_sf()
```

```{r}
country_boundary_US <- st_read("data/NEON-DS-Site-Layout-Files/US-Boundary-Layers/US-Boundary-Dissolved-States.shp")
```

```{r}
ggplot() +
  geom_sf(data = country_boundary_US, color = "gray18", size = 2) +
  geom_sf(data = state_boundary_US, color = "gray40") +
  ggtitle("Map of Contiguous US State Boundaries") +
  coord_sf()
```

```{r}
st_crs(point_HARV)
```

```{r}
st_crs(state_boundary_US)
```

```{r}
st_crs(country_boundary_US)
```

```{r}
st_bbox(point_HARV)
```

```{r}
st_bbox(state_boundary_US)
```

```{r}
ggplot() +
  geom_sf(data = country_boundary_US, size = 2, color = "gray18") +
  geom_sf(data = state_boundary_US, color = "gray40") +
  geom_sf(data = point_HARV, shape = 19, color = "purple") +
  ggtitle("Map of Contiguous US State Boundaries") +
  coord_sf()
```

```{r}
NE.States.Boundary.US <- st_read("data/NEON-DS-Site-Layout-Files/US-Boundary-Layers/Boundary-US-State-NEast.shp")
```

```{r}
ggplot() +
    geom_sf(data = NE.States.Boundary.US, aes(color ="color"), show.legend = "line") +
    scale_color_manual(name = "", labels = "State Boundary", values = c("color" = "gray18")) +
    geom_sf(data = point_HARV, aes(shape = "shape"), color = "purple") +
    scale_shape_manual(name = "", labels = "Fisher Tower", values = c("shape" = 19)) +
    ggtitle("Fisher Tower location") +
    theme(legend.background = element_rect(color = NA)) +
    coord_sf()
```

# Convert from .csv to a Shapefile (40 + 20 minutes)

## Import .csv
```{r}
plot_locations_HARV <-
  read.csv("data/NEON-DS-Site-Layout-Files/HARV/HARV_PlotLocations.csv")

str(plot_locations_HARV)
```

```{r}
names(plot_locations_HARV)
```

```{r}
head(plot_locations_HARV$easting)
```

```{r}
head(plot_locations_HARV$northing)
```

```{r}
head(plot_locations_HARV$geodeticDa)
```

```{r}
head(plot_locations_HARV$utmZone)
```

```{r}
st_crs(point_HARV)
```

```{r}
utm18nCRS <- st_crs(point_HARV)
utm18nCRS
```

```{r}
class(utm18nCRS)
```

```{r}
plot_locations_sp_HARV <- st_as_sf(plot_locations_HARV, coords = c("easting", "northing"), crs = utm18nCRS)
```

```{r}
st_crs(plot_locations_sp_HARV)
```

```{r}
ggplot() +
  geom_sf(data = plot_locations_sp_HARV) +
  ggtitle("Map of Plot Locations")
```

```{r}
ggplot() +
  geom_sf(data = aoi_boundary_HARV) +
  geom_sf(data = plot_locations_sp_HARV) +
  ggtitle("AOI Boundary Plot")
```

### Challenge

```{r}
newplot_locations_HARV <-
  read.csv("data/NEON-DS-Site-Layout-Files/HARV/HARV_2NewPhenPlots.csv")
str(newplot_locations_HARV)
```

```{r}
geogCRS <- st_crs(country_boundary_US)
geogCRS
```

```{r}
newPlot.Sp.HARV <- st_as_sf(newplot_locations_HARV, coords = c("decimalLon", "decimalLat"), crs = geogCRS)
```

```{r}
st_crs(newPlot.Sp.HARV)
```

```{r}
ggplot() +
  geom_sf(data = plot_locations_sp_HARV, color = "orange") +
  geom_sf(data = newPlot.Sp.HARV, color = "lightblue") +
  ggtitle("Map of All Plot Locations")
```

## Export a shapefile
```{r}
st_write(plot_locations_sp_HARV,
         "data/PlotLocations_HARV.shp", driver = "ESRI Shapefile")
```
